---
title: "Final report"
output: html_document
date: '2022-04-28'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section 1: Loading the data set and libraries

Main dataset - data  
Copy of the dataset used for cleaning - datacopy  
Final cleaned dataset - df  

### Loading the libraries
```{r message = FALSE, warnings = FALSE}
# loading the libraries
library(funModeling)
library(dplyr)
library(ggplot2)
library(ggmosaic)
library(plotrix)
library(chron)
library(corrplot)

```
### Loading the file

```{r}

data <- read.csv("37-00049_UOF-P_2016_prepped.csv")
```

## Section 2: Cleaning the dataset  
## Section 3: Exploring the data set after cleaning  
## Section 4: Exploratory data analysis  
## Section 5: Geo spatial mapping  
## Section 6: interpretations  
## Section 7: Conclusion  

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Trial

Adding new stuff here

```
{}
```